Safe Redis Queries

### 2016-06-12 aplas16/sections/IndexedMonads.tex
Popcorn -> Edis.

We define a new indexed monad \ensuremath{\Conid{Popcorn}} which, at term level, merely wraps
\ensuremath{\Conid{Redis}} in an additional constructor. The purpose is to add the
pre/postconditions at type level:
\begin{hscode}\SaveRestoreHook
\column{B}{@{}>{\hspre}l<{\hspost}@{}}%
\column{5}{@{}>{\hspre}l<{\hspost}@{}}%
\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
\>[B]{}\mathbf{newtype}\;\Conid{Popcorn}\;\Varid{p}\;\Varid{q}\;\Varid{a}\mathrel{=}\Conid{Popcorn}\;\{\mskip1.5mu \Varid{unPopcorn}\mathbin{::}\Conid{Redis}\;\Varid{a}\mskip1.5mu\}~~,{}\<[E]%
\\[\blanklineskip]%
\>[B]{}\mathbf{instance}\;\Conid{IMonad}\;\Conid{Popcorn}\;\mathbf{where}{}\<[E]%
\\
\>[B]{}\hsindent{5}{}\<[5]%
\>[5]{}\Varid{unit}\mathrel{=}\Conid{Popcorn}\mathbin{\cdot}\Varid{return}{}\<[E]%
\\
\>[B]{}\hsindent{5}{}\<[5]%
\>[5]{}\Varid{bind}\;\Varid{m}\;\Varid{f}\mathrel{=}\Conid{Popcorn}\;(\Varid{unPopcorn}\;\Varid{m}\bind \Varid{unPopcorn}\mathbin{\cdot}\Varid{f})~~.{}\<[E]%
\ColumnHook
\end{hscode}\resethooks
At term level, the \ensuremath{\Varid{unit}} and \ensuremath{\Varid{bind}} methods are not interesting: they merely
make calls to \ensuremath{\Varid{return}} and \ensuremath{(\bind )} of \ensuremath{\Conid{Redis}}, and extracts and re-apply the constructor \ensuremath{\Conid{Popcorn}} when necessary. With \ensuremath{\Conid{Popcorn}} being a \ensuremath{\mathbf{newtype}}, they
can be optimized away in runtime. The interesting bits happen in compile type,
on the added type information.

@@

The command \text{PING} in \Redis{} does nothing but replies a message
\text{PONG} if the connection is alive. In \Hedis{}, \ensuremath{\Varid{ping}} has type
\ensuremath{\Conid{Redis}\;(\Conid{Either}\;\Conid{Reply}\;\Conid{Status})}. The \Popcorn{} version of \ensuremath{\Varid{ping}} simply
applys an additional constructor (functions from \Hedis{} are qualified with
\ensuremath{\Conid{Hedis}} to prevent name clashing):
\begin{hscode}\SaveRestoreHook
\column{B}{@{}>{\hspre}l<{\hspost}@{}}%
\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
\>[B]{}\Varid{ping}\mathbin{::}\Conid{Popcorn}\;\Varid{xs}\;\Varid{xs}\;(\Conid{Either}\;\Conid{Reply}\;\Conid{Status}){}\<[E]%
\\
\>[B]{}\Varid{ping}\mathrel{=}\Conid{Popcorn}\;\Varid{\Conid{Hedis}.ping}~~.{}\<[E]%
\ColumnHook
\end{hscode}\resethooks
Since \ensuremath{\Varid{ping}} does not alter the database, the postcondition and precondition
are the same. Commands that are more interesting will be introduced after
we present our type-level encoding of constraints on states.

>>>>

We define a new indexed monad \ensuremath{\Conid{Edis}} which, at term level, merely wraps
\ensuremath{\Conid{Redis}} in an additional constructor. The purpose is to add the
pre/postconditions at type level:
\begin{hscode}\SaveRestoreHook
\column{B}{@{}>{\hspre}l<{\hspost}@{}}%
\column{5}{@{}>{\hspre}l<{\hspost}@{}}%
\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
\>[B]{}\mathbf{newtype}\;\Conid{Edis}\;\Varid{p}\;\Varid{q}\;\Varid{a}\mathrel{=}\Conid{Edis}\;\{\mskip1.5mu \Varid{unEdis}\mathbin{::}\Conid{Redis}\;\Varid{a}\mskip1.5mu\}~~,{}\<[E]%
\\[\blanklineskip]%
\>[B]{}\mathbf{instance}\;\Conid{IMonad}\;\Conid{Edis}\;\mathbf{where}{}\<[E]%
\\
\>[B]{}\hsindent{5}{}\<[5]%
\>[5]{}\Varid{unit}\mathrel{=}\Conid{Edis}\mathbin{\cdot}\Varid{return}{}\<[E]%
\\
\>[B]{}\hsindent{5}{}\<[5]%
\>[5]{}\Varid{bind}\;\Varid{m}\;\Varid{f}\mathrel{=}\Conid{Edis}\;(\Varid{unEdis}\;\Varid{m}\bind \Varid{unEdis}\mathbin{\cdot}\Varid{f})~~.{}\<[E]%
\ColumnHook
\end{hscode}\resethooks
At term level, the \ensuremath{\Varid{unit}} and \ensuremath{\Varid{bind}} methods are not interesting: they merely
make calls to \ensuremath{\Varid{return}} and \ensuremath{(\bind )} of \ensuremath{\Conid{Redis}}, and extracts and re-apply the constructor \ensuremath{\Conid{Edis}} when necessary. With \ensuremath{\Conid{Edis}} being a \ensuremath{\mathbf{newtype}}, they
can be optimized away in runtime. The interesting bits happen in compile type,
on the added type information.

@@

The command \text{PING} in \Redis{} does nothing but replies a message
\text{PONG} if the connection is alive. In \Hedis{}, \ensuremath{\Varid{ping}} has type
\ensuremath{\Conid{Redis}\;(\Conid{Either}\;\Conid{Reply}\;\Conid{Status})}. The \Edis{} version of \ensuremath{\Varid{ping}} simply
applys an additional constructor (functions from \Hedis{} are qualified with
\ensuremath{\Conid{Hedis}} to prevent name clashing):
\begin{hscode}\SaveRestoreHook
\column{B}{@{}>{\hspre}l<{\hspost}@{}}%
\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
\>[B]{}\Varid{ping}\mathbin{::}\Conid{Edis}\;\Varid{xs}\;\Varid{xs}\;(\Conid{Either}\;\Conid{Reply}\;\Conid{Status}){}\<[E]%
\\
\>[B]{}\Varid{ping}\mathrel{=}\Conid{Edis}\;\Varid{\Conid{Hedis}.ping}~~.{}\<[E]%
\ColumnHook
\end{hscode}\resethooks
Since \ensuremath{\Varid{ping}} does not alter the database, the postcondition and precondition
are the same. Commands that are more interesting will be introduced after
we present our type-level encoding of constraints on states.

### 2016-06-12 aplas16/sections/Introduction.tex
Popcorn -> Edis.

This paper discusses the techniques we used and experiences we learned from building such a language, nicknamed \Popcorn{}. We constructed an {\em indexed
monad}, on top of the monad \ensuremath{\Conid{Redis}}, which is indexed by a dictionary that
maintains the set of currently defined keys and their types. To represent
the dictionary, we need to encode variable binds with {\em type-level} lists
and strings. To summarize our contributions:
\begin{itemize}
\item We present \Popcorn{}, a statically typed domain-specific language embedded in Haskell and built on \Hedis{}.
% also makes Redis polymorphic by automatically converting back and forth from values of arbitrary types and boring ByteStrings.
%
\item We demonstrate how to model variable bindings of an embedded DSL using
 language extensions including type-level literals and data kinds.
%
\item We provide (yet another) an example of encoding effects and constraints of
in types, with indexed monad~\cite{indexedmonad}, closed type-families~\cite{closedtypefamilies} and constraints kinds~\cite{constraintskinds}.
\end{itemize}
\todo{Phrase this better.}

>>>>

This paper discusses the techniques we used and experiences we learned from building such a language, nicknamed \Edis{}. We constructed an {\em indexed
monad}, on top of the monad \ensuremath{\Conid{Redis}}, which is indexed by a dictionary that
maintains the set of currently defined keys and their types. To represent
the dictionary, we need to encode variable binds with {\em type-level} lists
and strings. To summarize our contributions:
\begin{itemize}
\item We present \Edis{}, a statically typed domain-specific language embedded in Haskell and built on \Hedis{}.
% also makes Redis polymorphic by automatically converting back and forth from values of arbitrary types and boring ByteStrings.
%
\item We demonstrate how to model variable bindings of an embedded DSL using
 language extensions including type-level literals and data kinds.
%
\item We provide (yet another) an example of encoding effects and constraints of
in types, with indexed monad~\cite{indexedmonad}, closed type-families~\cite{closedtypefamilies} and constraints kinds~\cite{constraintskinds}.
\end{itemize}
\todo{Phrase this better.}

### 2016-06-12 aplas16/sections/TypeLevelDict.tex
Popcorn -> Edis.

The dictionary mixes values (strings such as \ensuremath{\text{\tt \char34 A\char34}}, \ensuremath{\text{\tt \char34 B\char34}}) and types. Further
more, as mentioned in Section~\ref{sec:indexed-monads}, the dictionaries are
to be used parameters to the indexed monad \ensuremath{\Conid{Popcorn}}. In a dependently typed
programming language (without the so-called ``phase distinction'' ---
separation between types and terms), this would pose no problem. In Haskell
however, the dictionaries, to index a monad, has to be a type as well.

@@

The \Hedis{} function \ensuremath{\Varid{del}\mathbin{::}[\mskip1.5mu \Conid{ByteString}\mskip1.5mu]\to \Conid{Either}\;\Conid{Reply}\;\Conid{Integer}} takes a list
of keys (encoded to \ensuremath{\Conid{ByteString}}) and removes the entries having those keys in
the database. For simplicity, we consider creating a \Popcorn{} counterpart
that takes only one key. A first attempt may lead to something like the
following:
\begin{hscode}\SaveRestoreHook
\column{B}{@{}>{\hspre}l<{\hspost}@{}}%
\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
\>[B]{}\Varid{del}\mathbin{::}\Conid{String}\to \Conid{Popcorn}\;\Varid{xs}\;(\Conid{Del}\;\Varid{xs}\mathbin{?})\;(\Conid{Either}\;\Conid{Reply}\;\Conid{Integer}){}\<[E]%
\\
\>[B]{}\Varid{del}\;\Varid{key}\mathrel{=}\Conid{Popcorn}\mathbin{\$}\Varid{\Conid{Hedis}.del}\;[\mskip1.5mu \Varid{encodeKey}\;\Varid{key}\mskip1.5mu]~~.{}\<[E]%
\ColumnHook
\end{hscode}\resethooks
At term-level, our \ensuremath{\Varid{del}} merely calls \ensuremath{\Varid{\Conid{Hedis}.del}} (assuming that \ensuremath{\Varid{encodeKey}}
converts \ensuremath{\Varid{key}} to a \ensuremath{\Conid{ByteString}}). At type-level, if the status of the database
before \ensuremath{\Varid{del}} is called is specified by \ensuremath{\Varid{xs}}, the status after the execution
should be specified by \ensuremath{\Conid{Del}\;\Varid{xs}\mathbin{?}}. The question, however, is what to fill in
the question mark. The string \ensuremath{\Varid{key}} is a runtime value. How do we pass it to
the type level?

@@

\begin{hscode}\SaveRestoreHook
\column{B}{@{}>{\hspre}l<{\hspost}@{}}%
\column{5}{@{}>{\hspre}l<{\hspost}@{}}%
\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
\>[B]{}\Varid{del}\mathbin{::}\Conid{KnownSymbol}\;\Varid{s}{}\<[E]%
\\
\>[B]{}\hsindent{5}{}\<[5]%
\>[5]{}\Rightarrow \Conid{Proxy}\;\Varid{s}{}\<[E]%
\\
\>[B]{}\hsindent{5}{}\<[5]%
\>[5]{}\to \Conid{Popcorn}\;\Varid{xs}\;(\Conid{Del}\;\Varid{xs}\;\Varid{s})\;(\Conid{Either}\;\Conid{Reply}\;\Conid{Integer}){}\<[E]%
\\
\>[B]{}\Varid{del}\;\Varid{key}\mathrel{=}\Conid{Popcorn}\mathbin{\$}\Varid{\Conid{Hedis}.del}\;(\Varid{encodeKey}\;\Varid{key}){}\<[E]%
\ColumnHook
\end{hscode}\resethooks

>>>>

The dictionary mixes values (strings such as \ensuremath{\text{\tt \char34 A\char34}}, \ensuremath{\text{\tt \char34 B\char34}}) and types. Further
more, as mentioned in Section~\ref{sec:indexed-monads}, the dictionaries are
to be used parameters to the indexed monad \ensuremath{\Conid{Edis}}. In a dependently typed
programming language (without the so-called ``phase distinction'' ---
separation between types and terms), this would pose no problem. In Haskell
however, the dictionaries, to index a monad, has to be a type as well.

@@

The \Hedis{} function \ensuremath{\Varid{del}\mathbin{::}[\mskip1.5mu \Conid{ByteString}\mskip1.5mu]\to \Conid{Either}\;\Conid{Reply}\;\Conid{Integer}} takes a list
of keys (encoded to \ensuremath{\Conid{ByteString}}) and removes the entries having those keys in
the database. For simplicity, we consider creating a \Edis{} counterpart
that takes only one key. A first attempt may lead to something like the
following:
\begin{hscode}\SaveRestoreHook
\column{B}{@{}>{\hspre}l<{\hspost}@{}}%
\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
\>[B]{}\Varid{del}\mathbin{::}\Conid{String}\to \Conid{Edis}\;\Varid{xs}\;(\Conid{Del}\;\Varid{xs}\mathbin{?})\;(\Conid{Either}\;\Conid{Reply}\;\Conid{Integer}){}\<[E]%
\\
\>[B]{}\Varid{del}\;\Varid{key}\mathrel{=}\Conid{Edis}\mathbin{\$}\Varid{\Conid{Hedis}.del}\;[\mskip1.5mu \Varid{encode}\;\Varid{key}\mskip1.5mu]~~.{}\<[E]%
\ColumnHook
\end{hscode}\resethooks
At term-level, our \ensuremath{\Varid{del}} merely calls \ensuremath{\Varid{\Conid{Hedis}.del}}, with the help of \ensuremath{\Varid{encode}}
that converts \ensuremath{\Varid{key}} to a \ensuremath{\Conid{ByteString}}. At type-level, if the status of the database before \ensuremath{\Varid{del}} is called is specified by \ensuremath{\Varid{xs}}, the status afterwards
should be specified by \ensuremath{\Conid{Del}\;\Varid{xs}\mathbin{?}}. The question, however, is what to fill in
the question mark. The string \ensuremath{\Varid{key}} is a runtime value. How do we pass it to
the type level?

@@

\begin{hscode}\SaveRestoreHook
\column{B}{@{}>{\hspre}l<{\hspost}@{}}%
\column{5}{@{}>{\hspre}l<{\hspost}@{}}%
\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
\>[B]{}\Varid{del}\mathbin{::}\Conid{KnownSymbol}\;\Varid{s}{}\<[E]%
\\
\>[B]{}\hsindent{5}{}\<[5]%
\>[5]{}\Rightarrow \Conid{Proxy}\;\Varid{s}{}\<[E]%
\\
\>[B]{}\hsindent{5}{}\<[5]%
\>[5]{}\to \Conid{Edis}\;\Varid{xs}\;(\Conid{Del}\;\Varid{xs}\;\Varid{s})\;(\Conid{Either}\;\Conid{Reply}\;\Conid{Integer}){}\<[E]%
\\
\>[B]{}\Varid{del}\;\Varid{key}\mathrel{=}\Conid{Edis}\mathbin{\$}\Varid{\Conid{Hedis}.del}\;(\Varid{encodeKey}\;\Varid{key}){}\<[E]%
\ColumnHook
\end{hscode}\resethooks
